---
title: "Tax Morale in Africa - the Black Box of Tax Compliance"
author: "Wiebke Weiger & Jasmin Cantzler"
date: "11 December 2015"
output: "html_document"
---

```{r, include=FALSE}
#setwd("C:/Users/Wiebke/Documents/RepRes/TaxMorale/Assignment 3/PreparatoryWork")
setwd("/Users/jasmincantzler/Documents/TaxMorale")

library(dplyr)
library(tidyr)
library(googleVis)
library(formattable)
library(ggplot2)

# Using the no factor one as we want to show the means in the map
mydata <- read.csv('Afrobarometer.nofactors.csv')

#creates a subset with only Tax Morale and country
summary_df <- mydata %>% 
  group_by(Country) %>% 
  summarise(
    meantaxmorale = mean(TaxMorale, na.rm=TRUE)
  )

TaxMorale_map <- gvisGeoChart(summary_df, locationvar = 'Country',
                      colorvar = 'meantaxmorale',
                      options = list(
                          colors = "['#e31b23', '#ffff00', '#00853f']",
                          region= '002'
                          ))

```

**Tax morale could play a vital role in enhancing country's ability to efficiently mobilize national resources needed to invest in development, relieve poverty and deliver public services. Despite its potential, tax morale remains the underdog in enhancing tax compliance; most studies treat ‘tax morale’ as a black box without discussing or even considering how it might arise or how it might be maintained. We opened the box and present our findings and suggestions to policy makers in this post.**

```{r, echo=FALSE, results='asis'}
print(TaxMorale_map, tag = 'chart')
```

**Tax morale is an issue impinging Africa's development**

High levels of tax evasion lead to a misallocation of resources and hinder the ability of the government to invest in the provision of public goods. Therefore, understanding the rationale behind tax evaders becomes essential to national development and state building. 

Traditionally, the standard way to explain tax compliance has been the economics-of-crime approach, which assumes a rational taxpayer maximizing his or her utility by balancing the risk of detection and punishment with the benefit of tax evasion. In recent years there has been a revelation that this approach is too narrow to fully explain tax compliance and with this recognition came the awareness for a need to include more social factors. 

These social factors constitute and influence the individual’s “intrinsic motivation to pay taxes”; i.e. their tax morale. We set out to investigate the determinants of tax morale in African countries and found that these are not much different from what determines tax morale in other regions, such as Europe or the United States.

**The determinants of tax morale seem largely independent of cultural environments**


**Policy makers should recognize the opportunity and implement polices that account for these insights**


**For further insights please see:**
insert link here to the pdf of the final report