---
title: "Tax Morale in Africa - the Black Box of Tax Compliance"
author: "Wiebke Weiger & Jasmin Cantzler"
date: "11 December 2015"
output: "html_document"
---

```{r, include=FALSE}
#setwd("C:/Users/Wiebke/Documents/RepRes/TaxMorale/Assignment 3/PreparatoryWork")
setwd("/Users/jasmincantzler/Documents/TaxMorale")

library(dplyr)
library(tidyr)
library(googleVis)
library(formattable)

# Using the no factor one as we want to show the means in the map
mydata <- read.csv('Afrobarometer.nofactors.csv')

#creates a subset with only Tax Morale and country
summary_df <- mydata %>% 
  group_by(Country) %>% 
  summarise(
    meantaxmorale = mean(TaxMorale, na.rm=TRUE)
  )

TaxMorale_map <- gvisGeoChart(summary_df, locationvar = 'Country',
                      colorvar = 'meantaxmorale',
                      options = list(
                          colors = "['#e31b23', '#ffff00', '#00853f']",
                          region= '002'
                          ))
Bar <- gvisBarChart(df)
plot(Bar)

```

**Tax morale could play a vital role in enhancing country's ability to efficiently mobilize national resources needed to invest in development, relieve poverty and deliver public services. Despite its potential, tax morale remains the underdog in enhancing tax compliance; most studies treat ‘tax morale’ as a black box without discussing or even considering how it might arise or how it might be maintained. We opened the box and present our findings and suggestions to policy makers in this post.**

```{r, echo=FALSE, results='asis'}
print(TaxMorale_map, tag = 'chart')
```




**For further insights please see:**
insert link here to the pdf of the final report