---
title: "Opening the Black Box - Determinants of Tax Morale in Africa"
author: "Wiebke Weiger & Jasmin Cantzler"
date: "13 November 2015"
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
---
\pagebreak

#Introduction

#Literature Review

#Method

##Design

##Data Selection

#Operationalization

#Analysis Procedure
As illustrated in the previous section, we have identified certain characteristics as determinants of the approach needed to make meaningful assertions concerning the relationship between our explanatory variable and our dependent variable. 

When a dependent variable has more than two categories and the values of each category have a
meaningful sequential order where a value is indeed ‘higher’ than the previous one, then you
can use ordinal logit. 

#Limitations and Delimitations


#Results and Discussion

#Conclusion



```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
