---
title: "Tax Morale in Africa"
author: "Wiebke Weiger & Jasmin Cantzler"
date: "22 November 2015"
output: 
  revealjs::revealjs_presentation:
    theme: sky
    transition: slide
    highlight: default
    center: true
    fig_caption: true
bibliography: References.bib
---
```{r, include=FALSE}
# Set working directory
#setwd("C:/Users/Wiebke/Documents/RepRes/TaxMorale/Assignment 3/PreparatoryWork")
setwd("/Users/jasmincantzler/Documents/TaxMorale/Final Project")

library(dplyr)
library(tidyr)
library(googleVis)

# Using the no factor one as we want to show the means in the map
mydata <- read.csv('Afrobarometer.nofactors.csv')

#creates a subset with only Tax Morale and country
summary_df <- mydata %>% 
  group_by(Country) %>% 
  summarise(
    meantaxmorale = mean(TaxMorale, na.rm=TRUE)
  )

TaxMorale_map <- gvisGeoChart(summary_df, locationvar = 'Country',
                      colorvar = 'meantaxmorale',
                      options = list(
                          colors = "['#e31b23', '#ffff00', '#00853f']",
                          region= '002'
                          ))

```

## Why do people pay taxes?

The discourse about tax compliance was long dominated by Allingham & Sandmo's [-@Allingham1972] model of income tax evasion:

$$Compliance_i = \alpha_i + \beta_1 Probability of Detection_i \\
+ \beta_2 Rate of Penalty_i + \epsilon_i$$ 

But turns out their model had little explanatory power and failed to account for the high degree of tax compliance in some countries, despite their low level of deterrence.

## Tax compliance is “not only a function of opportunity, tax rates, probability of detection and so on...

but of each individual’s willingness to comply, shaped by tax morale” [@Torgler2007, 77]. 

Tax morale is often defined as **“the existence of an intrinsic motivation to pay taxes”** [@Cummings2009; @Torgler2005; @Torgler2009].

## But what determines Tax Morale?

$$Tax Morale_i = \alpha_i + \beta_1 Trust in Government_i \\
+ \beta_2 LevelCorruption_i + \beta_3 interactionstaxpayers_i \\ 
+ \beta_4 selfemployed_i + \beta_5 detection and punishment_i + \epsilon_i$$


## We investigated the determinants of tax morale in Africa, because...

understanding the rationale behind tax evaders becomes essential to national development and state building.

## Tax Morale varies by country...

```{r, echo=FALSE, results='asis', fig.align: ('center'), fig.cap="Tax Morale in Africa"}
print(TaxMorale_map, tag = 'chart')
```

## Determinants of Tax Morale however are quite similar...


## References
