---
title: "Tax Morale in Africa"
author: "Jasmin Cantzler & Wiebke Weiger"
date: "4 December 2015"
output: 
  revealjs::revealjs_presentation:
    theme: sky
    transition: none
    slideNumber: 'c/t'
    highlight: default
    center: true
    fig_caption: true
bibliography: References.bib
---
```{r, include=FALSE}
# Set working directory
setwd("C:/Users/Wiebke/Documents/RepRes/TaxMorale/Final Project")
#setwd("/Users/jasmincantzler/Documents/TaxMorale/Final Project")

library(knitr)
library(dplyr)
library(tidyr)
library(googleVis)
library(revealjs)
library(stringi)
library(ggplot2)
library(MASS)
library(stargazer)



# Using the no factor one as we want to show the means in the map
data.numeric <- read.csv('Afrobarometer.nofactors.csv')
data.factors <- read.csv('Afrobarometer.final.csv')

#creates a subset with only Tax Morale and country
summary_df <- data.numeric %>% 
  group_by(Country) %>% 
  summarise(
    meantaxmorale = mean(TaxMorale, na.rm=TRUE)
  )

TaxMorale_map <- gvisGeoChart(summary_df, locationvar = 'Country',
                      colorvar = 'meantaxmorale',
                      options = list(
                          colors = "['#7f7fff', '#000019']",
                          region= '002',
                          width=900, height=700
                          ))
plot(TaxMorale_map)


# Histogram of distribution of dependent variable tax morale:

data.factors$TaxMorale = factor(data.factors$TaxMorale,levels(data.factors$TaxMorale)[c(5,2,3,1,4)])

levels(data.factors$TaxMorale)

taxmorale_dist <- qplot(data.factors$TaxMorale,
      geom="histogram",
      binwidth=6,
      main="Distribution of responses",
      xlab="The tax department always has the right to make people pay taxes.",
      fill=I("lightblue"))

print(taxmorale_dist, tag = 'chart')

# Histogram of distribution of dependent variable tax morale (only 3 levels):

#data.3levels <- data.numeric

#data.3levels$TaxMorale[data.3levels$TaxMorale==2] <- 1
#data.3levels$TaxMorale[data.3levels$TaxMorale==3] <- 2
#data.3levels$TaxMorale[data.3levels$TaxMorale==4] <- 3
#data.3levels$TaxMorale[data.3levels$TaxMorale==5] <- 3


#data.3levels$TaxMorale <- cut(data.3levels$TaxMorale,
#                            breaks = c(0, 1, 2, 3),
#                            labels = c("Disagree", "Neither agree nor disagre", "Agree"))

#levels(data.3levels$TaxMorale)

#taxmorale_dist3levels <- qplot(data.3levels$TaxMorale,
#      geom="histogram",
#      binwidth=6,
#      main="Distribution of responses",
#      xlab="The tax department always has the right to make people pay taxes.",
#      fill=I("lightblue"))

#print(taxmorale_dist3levels, tag = 'chart')

##########################################################################################################
## Some preparations:
# First: Sort factor levles
# Second: Turn "Year" variable into factor
# Third: Recode dependent variable "TaxMorale" so that it has 3 instead of 5 levels

# First: Sort factor levels (they are currently in alphabetical order) for output to be more easily
# readable for the independent Variables:

# A) TRUST:
# 1) Trust key leadership figure (President/Prime Minister)
levels(data.factors$TrustPresident) # gives the current order of the levels:
# [1] "A lot"         "Just a little" "Not at all"    "Somewhat"
# correct order would be: 3, 2, 4, 1
# change order: 
data.factors$TrustPresident = factor(data.factors$TrustPresident,levels(data.factors$TrustPresident)[c(3,2,4,1)])
levels(data.factors$TrustPresident) # now order is correct:
# [1] "Not at all"    "Just a little" "Somewhat"      "A lot" 



# B) CORRUPTION:
# 1) Office of the President/Prime Minister involved in corruptoion
levels(data.factors$CorruptionPresident) # gives the current order of the levels:
# [1] "All of them"  "Most of them" "None"         "Some of them"
# correct order would be: 3, 4, 2, 1 
# change order: 
data.factors$CorruptionPresident = factor(data.factors$CorruptionPresident,levels(data.factors$CorruptionPresident)[c(3,4,2,1)])
levels(data.factors$CorruptionPresident) # now order is correct:
# [1] "None"         "Some of them" "Most of them" "All of them"

# C) Living conditions
levels(data.factors$LivingConditions) # gives the current order of the levels:
# [1] "Fairly bad"           "Fairly good"          "Neither good nor bad" "Very bad"            
# [5] "Very good"
# correct order would be: 4,1,3,2,5 
# change order: 
data.factors$LivingConditions = factor(data.factors$LivingConditions,levels(data.factors$LivingConditions)[c(4,1,3,2,5)])
levels(data.factors$LivingConditions) # now order is correct:
# [1] "Very bad"             "Fairly bad"           "Neither good nor bad" "Fairly good"         
# [5] "Very good"

# D) Gender
data.factors$Gender = factor(data.factors$Gender,levels(data.factors$Gender)[c(2,1)])

# E) Religious
levels(data.factors$Religion) # gives the current order of the levels:
# [1] "Active member"   "Inactive member" "Not a member"    "Official Leader"
# correct order would be: 3,2,1,4 
# change order: 
data.factors$Religion = factor(data.factors$Religion,levels(data.factors$Religion)[c(3,2,1,4)])
levels(data.factors$Religion) # now order is correct:
# [1] "Not a member"    "Inactive member" "Active member"   "Official Leader"




# Second: turn "Year"-variable into factor:

str(data.factors) # shows that Year is currently treated as a numeric variable
data.factors$Year <- cut(data.factors$Year,
                              breaks = c(0, 2005, 2008, 2015),
                              labels = c("2005", "2008", "2015"))
str(data.factors) # shows that Year is now a factor vairable with 3 levels
levels(data.factors$Year) # in correct order: 2005, 2008, 2015



# Third: Recode dependent variable "TaxMorale" so that it has 3 instead of 5 levels

#data.3levels <- data.numeric

#data.3levels$TaxMorale[data.3levels$TaxMorale==2] <- 1
#data.3levels$TaxMorale[data.3levels$TaxMorale==3] <- 2
#data.3levels$TaxMorale[data.3levels$TaxMorale==4] <- 3
#data.3levels$TaxMorale[data.3levels$TaxMorale==5] <- 3

#data.3levels$TaxMorale <- cut(data.3levels$TaxMorale,
#                              breaks = c(0, 1, 2, 3),
#                              labels = c("Disagree", "Neither agree nor disagre", "Agree"))

#levels(data.3levels$TaxMorale) 

# end of preparatory work
########################################################################################################

# MODELS

# 1. Models including data for all three years (only those variables that are available in all rounds)

# Model 1.1: 
# Tax morale as a function of 
#   - trust in the government (trying out various versions, e.g. trust in president, trust in parliament, etc)
#   - level of corruption (trying out various versions, e.g. president corrupt, parliament corrupt, etc.)
#   - country fixed effects
#   - year fixed effects

reg1.1 <- polr(TaxMorale ~ TrustPresident + CorruptionPresident + Year + Country, method='logistic', 
               data=data.factors, Hess = TRUE)

summary(reg1.1) ## 26845 observations deleted due to missingness (that is 25.6% of all observations)


# Store table
(ctable1.1 <- coef(summary(reg1.1)))

# Calculate and store p values
p1.1 <- pnorm(abs(ctable1.1[, "t value"]), lower.tail = FALSE) * 2

# Combined table
(ctable1.1 <- cbind(ctable1.1, "p value" = p1.1))

# Confidence Interval
ci.1.1 <- confint.default(reg1.1)

# Odds Ratios
exp(coef(reg1.1))

# Combine Odds Ratios and Confidence Interval
ORtable <- exp(cbind(OR = coef(reg1.1), ci.1.1))

kable(ORtable)



# Model 1.2 (including socio-economic controls):
#   - trust in the government (trying out various versions, e.g. trust in president, trust in parliament, etc)
#   - level of corruption (trying out various versions, e.g. president corrupt, parliament corrupt, etc.)
#   - country fixed effects
#   - year fixed effects
#   - living conditions
#   - age
#   - gender
#   - religious 

reg1.2 <- polr(TaxMorale ~ TrustPresident + CorruptionPresident +  Year + Country + 
               LivingConditions + Age + Gender + Religion, method='logistic', 
               data=data.factors, Hess = TRUE)

summary(reg1.2) ## 27912 observations deleted due to missingness (that is 26.7% of all observations)

# Store table
(ctable1.2 <- coef(summary(reg1.2)))

# Calculate and store p values
p1.2 <- pnorm(abs(ctable1.2[, "t value"]), lower.tail = FALSE) * 2

# Combined table
(ctable1.2 <- cbind(ctable1.2, "p value" = p1.2))

# Confidence Interval
ci.1.2 <- confint.default(reg1.2)

# Odds Ratios
exp(coef(reg1.2))

# Combine Odds Ratios and Confidence Interval
exp(cbind(OR = coef(reg1.2), ci.1.2))



#####
# Joint table for Models 1.1 and 1.2

library(stargazer)
table <- stargazer(reg1.1, reg1.2,
          title = 'Ordinal Logistic Regression Results of Tax Morale',
          digits = 2, type = 'html')




```

## Why do people pay taxes?

The discourse about tax compliance was long dominated by Allingham & Sandmo's (1972) model of income tax evasion:

$$Compliance_i = \alpha_i + \beta_1 ProbabilityOfDetection_i \\
+ \beta_2 RateOfPenalty_i + \epsilon_i$$ 

BUT: Their model had little explanatory power and failed to account for cases of high degrees of tax compliance despite low levels of deterrence.

## Tax compliance is "not only a function of...

- opportunity
- tax rates
- probability of detection 
- and so on...
- but of each individual's willingness to comply, shaped by tax morale" (Torgler 2007: 77) 

Tax morale is often defined as 
**"the existence of an intrinsic motivation to pay taxes"** 

## What determines Tax Morale?

$$Tax Morale_i = \alpha_i + \beta_1 TrustInGovernment_i \\
+ \beta_2 LevelCorruption_i + \beta_3 InteractionTaxpayers_i \\ 
+ \beta_4 Selfemployed_i + \beta_5 DetectionAndPunishment_i + \epsilon_i$$


## Research Question and Motivation

- What are the determinants of tax morale in Africa?
- Existing research focuses on developed countries
- Understanding the drivers of tax morale particularly important for developing countries
- **Do identified determinants of tax morale (defined mostly in the US or the EU) also hold in African countries?** 

## Our Data...

- ...comes from the [Afrobarometer Survey](http://www.afrobarometer.org/) 
- ...is individual level survey data (no panel)
- ...was merged for three different survey rounds:
  
    + Round 5: 34 countries (2011-2013)
 
    + Round 4: 20 countries (2008-2010)
 
    + Round 3: **18 countries** (2005-2006)
    
- ...contains a total of 104,697 observations 


## Operationalization of Tax Morale

People were asked whether they agree or disagree with the statement

**The tax authorities always have the right to make people pay taxes.**

They could choose between:

  - strongly disagree
  - disagree
  - neither agree nor disagree
  - agree
  - strongly agree
  

## To get a first idea:

```{r, echo=FALSE, results='asis'}
print(taxmorale_dist, tag = 'chart')
```
 

## Average Tax Morale in Africa


```{r, echo=FALSE, results='asis', fig.cap="Tax Morale in Africa"}
print(TaxMorale_map, tag = 'chart')
```


## Our Model to Explain Tax Morale

ordered logistic regression

**Control variables:**

  - trust in government and institutions
  - perceived corruption 
  - perceived level of tax avoidance
  - living conditions
  - self employement 
  - religion
  - age & gender
  - year & country fixed effects



## Results

```{r, echo=FALSE, results='asis', tidy=TRUE}
library(stargazer)
stargazer(reg1.1, reg1.2,
          title = 'Ordinal Logistic Regression Results of Tax Morale',
          digits = 2, type = 'html')
```

## Results (Cont.)
 
```{r, echo=FALSE, results='asis', tidy=TRUE}
library(knitr)
ortable <- kable(ORtable, align = 'c', digits = 2)
print(ortable)
```


## Shortcomings, Open Questions & Outlook for Future Research
- survey answers could be biased (social desirability)
- address model assumption of ordinal odds
- take a closer look at individual countries and between-country-differences (potentially evaluating impacts of campaigns etc.)


## Thank you very much for listening
