---
title: "Tax Morale in Africa"
author: "Jasmin Cantzler & Wiebke Weiger"
date: "4 December 2015"
output: 
  revealjs::revealjs_presentation:
    theme: sky
    transition: none
    slideNumber: 'c/t'
    highlight: default
    center: true
    fig_caption: true
bibliography: References.bib
---
```{r, include=FALSE}
# Set working directory
setwd("C:/Users/Wiebke/Documents/RepRes/TaxMorale/Final Project")
#setwd("/Users/jasmincantzler/Documents/TaxMorale/Final Project")

library(dplyr)
library(tidyr)
library(googleVis)
library(revealjs)
library(stringi)
library(ggplot2)
library(stargazer)


# Using the no factor one as we want to show the means in the map
data.numeric <- read.csv('Afrobarometer.nofactors.csv')
data.factors <- read.csv('Afrobarometer.final.csv')

#creates a subset with only Tax Morale and country
summary_df <- data.numeric %>% 
  group_by(Country) %>% 
  summarise(
    meantaxmorale = mean(TaxMorale, na.rm=TRUE)
  )

TaxMorale_map <- gvisGeoChart(summary_df, locationvar = 'Country',
                      colorvar = 'meantaxmorale',
                      options = list(
                          colors = "['#7f7fff', '#000019']",
                          region= '002',
                          width=900, height=700
                          ))
plot(TaxMorale_map)


# Histogram of distribution of dependent variable tax morale:

data.factors$TaxMorale = factor(data.factors$TaxMorale,levels(data.factors$TaxMorale)[c(5,2,3,1,4)])

levels(data.factors$TaxMorale)

taxmorale_dist <- qplot(data.factors$TaxMorale,
      geom="histogram",
      binwidth=6,
      main="Distribution of responses",
      xlab="The tax department always has the right to make people pay taxes.",
      fill=I("lightblue"))

print(taxmorale_dist, tag = 'chart')

# Histogram of distribution of dependent variable tax morale (only 3 levels):

data.3levels <- data.numeric

data.3levels$TaxMorale[data.3levels$TaxMorale==2] <- 1
data.3levels$TaxMorale[data.3levels$TaxMorale==3] <- 2
data.3levels$TaxMorale[data.3levels$TaxMorale==4] <- 3
data.3levels$TaxMorale[data.3levels$TaxMorale==5] <- 3


data.3levels$TaxMorale <- cut(data.3levels$TaxMorale,
                            breaks = c(0, 1, 2, 3),
                            labels = c("Disagree", "Neither agree nor disagre", "Agree"))

levels(data.3levels$TaxMorale)

taxmorale_dist3levels <- qplot(data.3levels$TaxMorale,
      geom="histogram",
      binwidth=6,
      main="Distribution of responses",
      xlab="The tax department always has the right to make people pay taxes.",
      fill=I("lightblue"))

print(taxmorale_dist3levels, tag = 'chart')

```

## Why do people pay taxes?

The discourse about tax compliance was long dominated by Allingham & Sandmo's (1972) model of income tax evasion:

$$Compliance_i = \alpha_i + \beta_1 ProbabilityOfDetection_i \\
+ \beta_2 RateOfPenalty_i + \epsilon_i$$ 

BUT: Their model had little explanatory power and failed to account for cases of high degrees of tax compliance despite low levels of deterrence.

## Tax compliance is "not only a function of...

- opportunity
- tax rates
- probability of detection 
- and so on...
- but of each individual's willingness to comply, shaped by tax morale" (Torgler 2007: 77) 

Tax morale is often defined as 
**"the existence of an intrinsic motivation to pay taxes"** 

## What determines Tax Morale?

$$Tax Morale_i = \alpha_i + \beta_1 TrustInGovernment_i \\
+ \beta_2 LevelCorruption_i + \beta_3 InteractionTaxpayers_i \\ 
+ \beta_4 Selfemployed_i + \beta_5 DetectionAndPunishment_i + \epsilon_i$$


## Research Question and Motivation

- What are the determinants of tax morale in Africa?
- Existing research focuses on developed countries
- Understanding the drivers of tax morale particularly important for developing countries
- **Do identified determinants of tax morale (defined mostly in the US or the EU) also hold in African countries?** 

## Our Data...

- ...comes from the [Afrobarometer Survey](http://www.afrobarometer.org/) 
- ...is individual level survey data (no panel)
- ...was merged for three different survey rounds:
  
    + Round 5: 34 countries (2011-2013)
 
    + Round 4: 20 countries (2008-2010)
 
    + Round 3: **18 countries** (2005-2006)
    
- ...contains a total of 104,697 observations 


## Operationalization of Tax Morale

People were asked whether they agree or disagree with the statement

**The tax authorities always have the right to make people pay taxes.**

They could choose between:

  - strongly disagree
  - disagree
  - neither agree nor disagree
  - agree
  - strongly agree
  

## To get a first idea:

```{r, echo=FALSE, results='asis'}
print(taxmorale_dist, tag = 'chart')
```
 

## Average Tax Morale in Africa


```{r, echo=FALSE, results='asis', fig.cap="Tax Morale in Africa"}
print(TaxMorale_map, tag = 'chart')
```


## Our Model to Explain Tax Morale

ordered logistic regression

**Control variables:**

  - trust in government and institutions
  - perceived corruption 
  - perceived level of tax avoidance
  - living conditions
  - self employement 
  - religion
  - age & gender
  - year & country fixed effects



## Results



## Shortcomings, Open Questions & Outlook for Future Research
- survey answers could be biased (social desirability)
- address model assumption of ordinal odds
- 
