---
title: "Tax Morale in Africa"
author: "Wiebke Weiger & Jasmin Cantzler"
date: "13 November 2015"
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
---
\pagebreak
```{r, include=FALSE}
##################################
# Assignment 3
# Jasmin Cantzler & Wiebke Weiger
# 13 November 2015
##################################

# Set working directory
#setwd("C:/Users/Wiebke/Documents/RepRes/TaxMorale/Assignment 3/PreparatoryWork")
setwd("/Users/jasmincantzler/Documents/TaxMorale/Assignment 3/PreparatoryWork")

#STEP 1 OF ASSIGNMENT: GATHER WEB BASED DATA FROM AT LEAST TWO SOURCES, MERGE THE DATA SETS 
#to load the R file that contains the merging of the different data sets please
# source('merging_data.R') 

# for sake of ease we directly load the final dataset that is the result of the file above
mydata <- read.csv('Afrobarometer.final.csv')

# Load packages
library(ggplot2)
library(knitr)
library(dplyr)

#creates Tax Morale as a numeric variable
TaxMoraleNumeric <- as.numeric(mydata$TaxMorale)

# CONDUCTS BASIC DESCRIPTIVE STATISTICS 
summary(mydata) 
summary(mydata$TaxMorale)

#creates a histogram of our dependent variable Tax Morale
qplot(mydata$TaxMorale,
      geom="histogram",
      binwidth=8,
      main="Summary Statistic",
      xlab="The tax department always has the right to make people pay taxes.",
      fill=I("lightblue"))

meanage <- mean(mydata$Age, na.rm=TRUE)
meanAge<- data.frame(meanage,mydata$Country)


table <- with(mydata,
              data.frame(Mean=mean(mydata$Age, na.rm = TRUE)),
              Country=factor(mydata$Country))
kable(table)

```

#Introduction

A country's ability to efficientyl mobilize its national resources is a topic that has recently been gaining traction in the discussion surrounding development. Since the Monterrey Consensus was adopted in 2002, in which countries recognized the importance of raising domestic revenues, development aid has been increasingly shifting from direct financial assistance to improving tax system and national resource management. This shift in strategy was officially stated in 2008, in the Doha Declaration on Financing for Development, which pledged to enhance national tax revenues; one of the recognized strategies pledged to achieve this was combating tax evasion.
  High levels of tax evasion lead to a misallocation of resources and hinder the ability of the government to invest in the provision of public goods. Therefore, understanding the rationale behind tax evaders becomes essential to national development and state building. Tradtionally, the standard way to explain tax compliance has been the economics-of-crime approach, which assumes a rational taxpayer maximizing his or her utility by balancing the risk of detection and punishment with the benefit of tax evasion. Today, this approach is increasingly regarded as too narrow to fully explain tax compliance, and many argue for the need to include social factors, which are said to explain why people conform to paying taxes even in the absence of strong deterrence mechanisms. These social factors constitute and influence the individual's "intrinsic motivation to pay taxes", hereafter referred to as tax morale, by increasing moral costs of tax evasion and thus increasing tax compliance.

Our research paper aims to shed light on the determinants of tax morale in African countries by conducting a regression analysis using data from the Afrobarometer for the years 2005-2006, 2008-2010 and 2011-2013. Our conceptual framework is based on commonly perceived determinants of tax morale in developed countries, whereby:

$$Tax Morale_i = \alpha_i + \beta_1 Trust in Government_i + $$

This paper briefly outlines the design, operationalization and data selection, before conducting the regression analysis and briefly describing its results. More extensive information will be available in the final report. 

#Method

##Data Selection and Operationalization

As outlined in Assignment 1, we make use of the Afrobarometer Surveys to investigate the determinants of tax morale in African countries. We merged the survey results from Round 3, 4 and 5, resulting in a data set that spans the years 2005-2006, 2008-2010 and 2011-2013.

##Descriptive Statistics of Our Data

Tax Morale is instrumented by a variable [^1] that asks the participants to agree or disagree with the statement "The tax department always has the right to make people pay taxes." The variable is measured from 1=Strongly Disagree, 2=Disagree, 3=Neither Agree Nor Disagree, 4=Agree, 5=Strongly Agree. 


**Table 1 in the columns 2, 3 and 4 shows for each country the percentage of individuals saying that tax evasion is never justifiable. Columns 5, 6, and 7 give the mean value for all countries based on a scale from 0 to 3, were 3 is the highest tax morale (value 0 integrates the values 4 to 10).**

```{r, echo=FALSE}
qplot(mydata$TaxMorale,
      geom="histogram",
      binwidth=8,
      main="Summary Statistic",
      xlab="The tax department always has the right to make people pay taxes.",
      fill=I("lightblue"))
```

The average values in Table 1 give a first overview about tax morale in African countries; in general .... seems to have higher tax morale than...

##Inferential Statistics

A regression analysis is used to investigate the determinants of tax morale in African countries:

$$Tax Morale_i = \alpha_i + \beta_1 Trust in Government_i + \beta_2 Level of Corruption_i + \beta_3 Interactions with other taxpayers_i + \beta_4 Tax Burden_i +  \beta_5 Social Norms_i + \epsilon_1$$

#Conclusion






[^1]: For Round 3 (Codebook 2005) it is variable Q52D, for Round 4 (Codebook 2008) it is Q44C, and for Round 5 (2015) it is Q48C.